Demo arXiv template

### 2022-11-21 paper/paper.qmd
first attempt to write notation into pipeline steps

---
title: "Demo arXiv template"
format:
  arxiv-pdf:
    keep-tex: true  
  arxiv-html: default
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, VIC
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@monash.edu
    url: https://huizezhangsh.netlify.app/
  - name: Collaborators
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, VIC
        country: Australia
abstract: |
  - indexes, useful, quantify severity, early monitoring, 
  - A huge number of indexes have been proposed by domain experts, however, a large majority of them are not being adopted, reused, and compared in research or in practice. 
  - One of the reasons for this is the plenty of indexes are quite complex and there is no obvious easy-to-use implementation to apply them to user's data. 
  - The paper describes a general pipeline framework to construct indexes from spatio-temporal data, 
  - This allows all the indexes to be constructed through a uniform data pipeline and different indexes to vary on the details of each step in the data pipeline and their orders. 
  - The pipeline proposed aim to smooth the workflow of index construction through breaking down the complicated steps proposed by various indexes into small building blocks shared by most of the indexes.
  - The framework will be demonstrated with drought indexes as examples, but appliable in general to environmental indexes constructed from multivariate spatio-temporal data
keywords:
  - indexes
  - data pipeline
  - software design
bibliography: bibliography.bib  
editor_options: 
  chunk_output_type: inline
---

@@

The construction of natural hazard indexes also follows a set of steps, which is usually illustrated using a flowchart in the paper.

* constructing time series index should also be encapsulated in the framework

Here we assume a concept of interest is determined, relevant variables/ indicators are identified and available to construct indexes.

Before the mathmetical construction of indexes

```{r}
#| fig-align: center
#| fig-cap: "Diagram of pipeline steps for index construction."
#| out-height: 90%
#| out-width: 100%
knitr::include_graphics(here::here("figures/pipeline-steps.png"))
```

@@

*Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem. After processing them and align them together they become the "raw data"*

The data used to construct the natural hazard index usually have three dimensions, one for location, one for time, and one for multivariate. 
Mathematically, it can be written as $X_{j, s, t}$, where $j = 1, 2, \cdots, J$ for variable, $s = 1, 2, \cdots, S$ for location, and $t = 1, 2, \cdots, T$ for time.

The location $s$ can refer to vector points or areas characterised by longitude-latitude coordinates, or raster cells obtained from satellite images.

The time dimension $t$ can be daily, weekly, biweekly (14-16 days), monthly, or even quarterly

Variables

This multidimensional array structure is commonly used in geospatial analysis

Given the variety of data sources at different spatial resolution and temporal granularity, the raw data may first come in multiple pieces. 
Sometimes, even a considerable amount of work is needed to align the spatial and temporal extent of multivariate data.

A notation for different variables have different spatial and temporal granularity $X_{j_1, s_1, t_1}$???

@@

mostly happen with raster data

@@

A specific transformation on the scale of the data

z-score standardising, min-max standardisation into [0, 1] or [0, 100], percentage change on the baseline
close to variable transformation step

@@

The purpose of normalising is for cross-comparison.
This step can get criticism from analysts for ...

specifically for converting from a fitted distribution to normal score via reverse CDF function, non-parametric formula, or empirical approximation, a common step in many index: SPI, SSI, Z score. The purpose of normalising is to convert the index into a standardised series after all the steps for the ease of comparison.

Normalising is usually the last step

<!-- Take one or multiple variables to create a new variable,  -->
Restrict it to single variable, square root, log etc could be linearly, also non-linear

change the shape of the variable

<!-- A conversion of non-normal variables into (approximately) normal. wrong! -->

GAM, can you do additive model pairwise/ three-way

sometimes called feature extraction in the machine learning community 
With drought indexes, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.]

Also include weighting

@@

@@

>>>>

---
title: "Demo arXiv template"
format:
  pdf:
    keep-tex: true  
    include-in-header: 
      text: |
        \usepackage{amsmath}
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, VIC
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@monash.edu
    url: https://huizezhangsh.netlify.app/
  - name: Collaborators
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, VIC
        country: Australia
abstract: |
  - indexes, useful, quantify severity, early monitoring, 
  - A huge number of indexes have been proposed by domain experts, however, a large majority of them are not being adopted, reused, and compared in research or in practice. 
  - One of the reasons for this is the plenty of indexes are quite complex and there is no obvious easy-to-use implementation to apply them to user's data. 
  - The paper describes a general pipeline framework to construct indexes from spatio-temporal data, 
  - This allows all the indexes to be constructed through a uniform data pipeline and different indexes to vary on the details of each step in the data pipeline and their orders. 
  - The pipeline proposed aim to smooth the workflow of index construction through breaking down the complicated steps proposed by various indexes into small building blocks shared by most of the indexes.
  - The framework will be demonstrated with drought indexes as examples, but appliable in general to environmental indexes constructed from multivariate spatio-temporal data
keywords:
  - indexes
  - data pipeline
  - software design
bibliography: bibliography.bib  
editor_options: 
  chunk_output_type: inline
---

@@

*constructing time series index should also be encapsulated in the framework*

@fig-pipeline-steps presents the index construction pipeline from the raw data to the final indexes. 
The available intermediate steps include temporal processing, spatial aggregation, variable transformation, scaling, dimension reduction, normalising, benchmarking, and simplification. 
These steps do not have to be arranged in sequence and not all the steps are required to present in every index construction. 
Spatial aggregation and temporal processing are operations of the data in space and time 
Variable transformation, scaling, and dimension reduction concerns operations on the multivariate aspect of the data.
Normalisation, benchmarking, and simplification can be viewed as the "post-processing" of indexes, which addresses the cross-comparison across different indexes and time, and index communication.

More steps are involved before the mathematical construction of indexes. 
This includes 1) defining a concept of interest, being useful for decision making and 2) finding the relevant and measurable variables to characterise the concept. 
These choices define the scope of the indexes and the data availability for constructing an index but not the construction itself, hence not included in the pipeline framework. 
The framework proposed in this paper assumes the objective of an index has been defined and relevant variables have been collected and quality controlled.

```{r fig-pipeline-steps}
#| fig-align: center
#| fig-cap: "Diagram of pipeline steps for index construction."
#| out-height: 90%
#| out-width: 100%
knitr::include_graphics(here::here("figures/pipeline-steps.png"))
```

@@

*Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\mathbf{s}$ and $\mathbf{t}$ to depend on $p$*

*Initially my thought was to raw data only after aligning the resolution: Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem. After processing them and align them together they become the "raw data"*

Once variables are collected, they can have different spatial resolution and temporal granularity. Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data that contain the spatial, temporal, and multivariate aspect: the spatial component $\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal component $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$. Expanding on the multivariate aspect, the data can be written as  $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$

<!-- The location $s$ can refer to vector points or areas characterised by longitude-latitude coordinates, or raster cells obtained from satellite images. -->

<!-- The time dimension $t$ can be daily, weekly, biweekly (14-16 days), monthly, or even quarterly -->

<!-- Variables -->

<!-- This multidimensional array structure is commonly used in geospatial analysis -->

<!-- Given the variety of data sources at different spatial resolution and temporal granularity, the raw data may first come in multiple pieces.  -->
<!-- Sometimes, even a considerable amount of work is needed to align the spatial and temporal extent of multivariate data. -->

Temporal processing concerns the operations on the temporal aspect of the data. It can be represented by $f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))$, where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.

- purpose is to aggregate across time, smooth?

An example of temporal processing is to aggregate the data, $x(s_i; t_j)$, into $k$-month sum. This can be used to reflect how the accumulated values change across time. In the drought index SPI, precipitation can be aggregated at different scales to reflect short-term, mid-term, and long-term drought severity. With $\psi = k$, the aggregated series can be written as $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)$ where $j^\prime$ is the time index after the aggregation.

@@

Sometimes, it can be beneficial to borrow information from nearby spatial locations or there is the need to combine raster data from satellite images with vector ground measures. Spatial aggregation can be written as $g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))$, where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.

 - $\theta$ can be a constant or a derived from the data

 - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level.

A variable transformation changes the shape of the variable. It can be written as $h_{\tau}(x(\mathbf{s};\mathbf{t}))$, where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the potential parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
Variable transformation involve square root transformation, log transformation, and also include those non-linear transformation.

<!-- Take one or multiple variables to create a new variable,  -->
<!-- Restrict it to single variable, square root, log etc could be linearly, also non-linear -->

<!-- change the shape of the variable -->

<!-- <!-- A conversion of non-normal variables into (approximately) normal. wrong! -->

<!-- GAM, can you do additive model pairwise/ three-way -->

@@

Scaling is a specific variable transformation on the scale of the data. 
The purpose to treat scaling as a separate step from variable transformation is to make the step explicit. 
Scaling is usually in the form of $[x(s_i;t_j) - \alpha]/\gamma$. For example, $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$ is used in a z-score standardisation to standardise the data by the its mean and standard deviation across all time and all space units and $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$ is used to standardise the variable into a unit interval.

<!-- z-score standardising, min-max standardisation into [0, 1] or [0, 100], percentage change on the baseline -->
<!-- close to variable transformation step -->

When multiple variables are collected in space and time to construct indices, they are often summarised with dimension reduction. Dimension reduction $x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})$,  where $p^* = 1, 2, \cdots, P^*$,  $p = 1, 2, \cdots, P$, and $P^* < P$. The most commonly used dimension reduction techniquePprincipal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF), can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the variance of the data given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.

<!-- sometimes called feature extraction in the machine learning community  -->
<!-- With drought indexes, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.] -->

@@

The purpose of normalising is for cross-comparison using the normal score.
This step can get criticism from analysts for ...

This step is usually in the end of the pipeline when the data have been processed into a univariate series. 
When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]$, where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$. 
Sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.

- In some cases, the distibution and inverse normal can be multivariate, i.e. use copula.

<!-- specifically for converting from a fitted distribution to normal score via reverse CDF function, non-parametric formula, or empirical approximation, a common step in many index: SPI, SSI, Z score. The purpose of normalising is to convert the index into a standardised series after all the steps for the ease of comparison. -->

@@

Benchmarking sets a constant value to allow the constructed index to be compared across time.
Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed, could be a constant or a function of the data, i.e. mean.

@@

In public communication, indexes are often delivered in categorical grades, along with its underlying numerical values. The simplification step prescribes how the continuous index values are converted into the discrete grades. This process can be written with the piece-wise function:

$$
\begin{cases}
C_0 & c_1 \leq (s_i; t_j) < c_0 \\
C_1 & c_2 \leq x(s_i; t_j) < c_1 \\
C_2 & c_3 \leq x(s_i; t_j) < c_2 \\
\cdots \\
C_z & c_z \leq x(s_i; t_j)
\end{cases}
$$

where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the threshold value in each category. In SPI, drought are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$. In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.
